---
title: "Cleaned_Up_Dom_Models"
output:
  pdf_document: default
  html_document: default
---

```{r, include = FALSE}
library(itscalledsoccer)
library(tidyverse)
library(mgcv)
library(lme4)
library(lmerTest)
library(lubridate)
library(broom)
library(broom.mixed)

asa_client <- AmericanSoccerAnalysis$new()

players <- asa_client$get_players(leagues = "mls")
teams   <- asa_client$get_teams(leagues = "mls")

seasons <- 2013:2025

### Build multi-season team panel

```


```{r, echo = FALSE}
# Team performance by season
asa_team_xgoals_panel <- asa_client$get_team_xgoals(
  leagues = "mls",
  season_name = seasons,
  split_by_seasons = TRUE
) |>
  left_join(teams, by = "team_id")

team_performance_panel <- asa_team_xgoals_panel |>
  mutate(season_name = as.integer(season_name)) |>
  group_by(team_id, team_name, season_name) |>
  summarize(
    xg_for = sum(xgoals_for, na.rm = TRUE),
    xg_against = sum(xgoals_against, na.rm = TRUE),
    xgd = xg_for - xg_against,
    .groups = "drop"
  )

# Multi-season salary data
mls_salaries_panel <- asa_client$get_player_salaries(
  leagues = "mls",
  season_name = seasons
) |>
  left_join(players, by = "player_id") |>
  mutate(
    season_name = as.integer(season_name),
    team_id = as.character(team_id)
  )

team_salary_panel <- mls_salaries_panel |>
  group_by(team_id, season_name) |>
  summarize(
    total_salary = sum(base_salary, na.rm = TRUE),
    .groups = "drop"
  )

panel_team_data <- team_performance_panel |>
  mutate(team_id = as.character(team_id)) |>
  left_join(team_salary_panel, by = c("team_id", "season_name")) |>
  mutate(
    log_salary = log(total_salary),
    season_name = factor(season_name)
  )
```


```{r, echo = FALSE}
### Baseline team model


panel_team_model <- lm(
  xgd ~ log_salary + season_name,
  data = panel_team_data
)

broom::tidy(panel_team_model, conf.int = TRUE)
```


```{r, echo = FALSE}
### Build player-season dataset

asa_player_gplus <- asa_client$get_player_goals_added(
  leagues = "mls",
  season_name = seasons,
  split_by_seasons = TRUE,
  split_by_teams = TRUE,
  above_replacement = TRUE
) |>
  mutate(
    season_name = as.integer(season_name),
    team_id = as.character(team_id)
  )

player_salaries_panel <- asa_client$get_player_salaries(
  leagues = "mls",
  season_name = seasons
) |>
  left_join(players, by = "player_id") |>
  mutate(
    season_name = as.integer(season_name),
    team_id = as.character(team_id)
  ) |>
  group_by(player_id, team_id, season_name) |>
  summarize(
    base_salary = mean(base_salary, na.rm = TRUE),
    position = first(position),
    .groups = "drop"
  )

player_season <- asa_player_gplus |>
  left_join(
    player_salaries_panel,
    by = c("player_id", "team_id", "season_name")
  ) |>
  left_join(players |> select(player_id, birth_date), by = "player_id") |>
  mutate(
    log_salary = log(base_salary + 1),
    position = factor(position),
    age = season_name - year(birth_date)
  ) |>
  filter(
    !is.na(goals_added_above_replacement),
    !is.na(log_salary),
    !is.na(position),
    position %in% c("D", "M", "F")
  ) |>
  mutate(
    position = droplevels(factor(position))
  )
```

```{r, echo = FALSE}
### Player-level GAM

fit_player_gam <- gam(
  goals_added_above_replacement ~
    s(log_salary, k = 5) +
    position +
    s(log_salary, by = position, k = 4),
  data = player_season,
  method = "REML"
)

broom::tidy(fit_player_gam, conf.int = TRUE)
```

```{r, echo = FALSE}
### Aggregate predicted player value to team-season

player_season_model <- player_season |>
  filter(position %in% c("D", "M", "F")) |>
  mutate(
    position = droplevels(factor(position))
  )

player_season_model <- player_season_model |>
  mutate(
    predicted_value = predict(fit_player_gam, newdata = player_season_model)
  )

team_efficiency <- player_season_model |>
  group_by(team_id, season_name) |>
  summarize(
    predicted_team_value = sum(predicted_value * minutes_played / 90, na.rm = TRUE),
    player_total_salary = sum(base_salary, na.rm = TRUE),
    avg_age = mean(age, na.rm = TRUE),
    pct_att = mean(position == "F"),
    pct_midfielders = mean(position == "M"),
    .groups = "drop"
  )
```

```{r, echo = FALSE}
### Main team model

team_efficiency_model_data <- panel_team_data |>
  mutate(
    team_id = as.character(team_id),
    season_name = as.integer(as.character(season_name))
  ) |>
  left_join(team_efficiency, by = c("team_id", "season_name")) |>
  filter(!is.na(xgd), !is.na(predicted_team_value), !is.na(player_total_salary))

team_efficiency_model <- lm(
  xgd ~ predicted_team_value + player_total_salary +
    factor(season_name) + avg_age + pct_att + pct_midfielders,
  data = team_efficiency_model_data
)

broom::tidy(team_efficiency_model, conf.int = TRUE)
```

```{r, echo = FALSE}

### Multilevel robustness model

team_efficiency_model_data <- team_efficiency_model_data |>
  mutate(
    pred_val_s = as.numeric(scale(predicted_team_value)),
    salary_s   = as.numeric(scale(player_total_salary)),
    age_s      = as.numeric(scale(avg_age)),
    pct_att_s  = as.numeric(scale(pct_att)),
    pct_mid_s  = as.numeric(scale(pct_midfielders))
  )

team_efficiency_model_ml <- lmer(
  xgd ~ pred_val_s + salary_s + age_s + pct_att_s + pct_mid_s +
    (1 | team_id) + (1 | season_name),
  data = team_efficiency_model_data
)

library(broom.mixed)

broom.mixed::tidy(team_efficiency_model_ml, effects = "fixed", conf.int = TRUE)
broom.mixed::tidy(team_efficiency_model_ml, effects = "ran_pars", conf.int = TRUE)
broom.mixed::tidy(team_efficiency_model_ml, effects = "ran_vals", conf.int = TRUE)
```

```{r, include = FALSE}
top3_tbl <- player_season |>
  group_by(team_id, season_name) |>
  arrange(desc(base_salary), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  summarize(
    top3_salary = sum(base_salary[rank <= 3], na.rm = TRUE),
    total_salary = sum(base_salary, na.rm = TRUE),
    top3_share = top3_salary / total_salary,
    .groups = "drop"
  )

u22_tbl <- player_season |>
  mutate(u22_proxy_conservative = age <= 22 & base_salary >= 200000 & base_salary <= 750000) |>
  group_by(team_id, season_name) |>
  summarize(
    u22_proxy_count = sum(u22_proxy_conservative, na.rm = TRUE),
    .groups = "drop"
  )

pos_tbl <- player_season |>
  group_by(team_id, season_name, position) |>
  summarize(
    pos_salary = sum(base_salary, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(team_id, season_name) |>
  mutate(
    total_salary = sum(pos_salary),
    pos_share = pos_salary / total_salary
  ) |>
  select(-pos_salary, -total_salary) |>
  filter(!is.na(position)) |>
  tidyr::pivot_wider(
    names_from = position,
    values_from = pos_share,
    values_fill = 0
  ) |>
  rename(
    share_D = D,
    share_M = M,
    share_F = F
  )

team_efficiency_model_data_ext <- team_efficiency_model_data |>
  left_join(top3_tbl, by = c("team_id", "season_name")) |>
  left_join(u22_tbl, by = c("team_id", "season_name")) |>
  left_join(pos_tbl, by = c("team_id", "season_name"))

team_efficiency_model_ml_ext <- lmer(
  xgd ~ pred_val_s + salary_s + age_s +
    top3_share + u22_proxy_count + share_F + share_M +
    (1 | team_id) + (1 | season_name),
  data = team_efficiency_model_data_ext
)

summary(team_efficiency_model_ml_ext)

team_efficiency_model_ml_ext2 <- lmer(
  xgd ~ pred_val_s + salary_s + age_s + top3_share + u22_proxy_count +
    share_F + share_M + (1 | team_id),
  data = team_efficiency_model_data_ext
)

summary(team_efficiency_model_ml_ext2)
```

```{r}
# Save Models
saveRDS(panel_team_model, "Models/panel_team_model.rds")
saveRDS(fit_player_gam, "Models/fit_player_gam.rds")
saveRDS(team_efficiency_model, "Models/team_efficiency_model.rds")
saveRDS(team_efficiency_model_ml, "Models/team_efficiency_model_ml.rds")
saveRDS(team_efficiency_model_ml_ext2, "Models/team_efficiency_model_ml_ext2.rds")

# also save final data objects if you need plots/tables later
saveRDS(panel_team_data, "Models/panel_team_data.rds")
saveRDS(player_season, "Models/player_season.rds")
saveRDS(team_efficiency_model_data, "Models/team_efficiency_model_data.rds")
saveRDS(team_efficiency_model_data_ext, "Data/team_efficiency_model_data_ext.rds")
saveRDS(mls_salaries_2025, "Data/mls_salaries_2025.rds")
saveRDS(team_model_data, "Data/team_model_data.rds")
saveRDS(position_salary, "Data/position_salary.rds")
```

---

## What each model does in your paper

Use this progression:

### 1. Baseline team model
Shows that raw spending alone has limited explanatory power.

### 2. Player GAM
Shows that salary translates into player value nonlinearly and differently by position.

### 3. Main team efficiency model
Shows that aggregated player value strongly predicts team performance, while salary is negative conditional on value.

### 4. Multilevel model
Shows the main result is robust after accounting for team- and season-level heterogeneity.

---

## What to report, model-wise

### Definitely report
- `panel_team_model`
- `fit_player_gam`
- `team_efficiency_model`
- `team_efficiency_model_ml`

### Do not report
- 2025-only raw salary / DP / U22 regressions
- interaction-heavy exploratory models
- early failed GAMs at team level

---

## Small writeup suggestion

In Methods/Results, frame the models as:

- **Baseline**
- **Mechanism**
- **Main team-level test**
- **Robustness check**


## EXTRA SHIT

#### Team Efficiency Model

$$
xGD_{it} = \beta_0 + \beta_1 \widehat{V}_{it} + \beta_2 \text{Salary}_{it} + \beta_3 \text{Age}_{it}
+ \beta_4 \text{PctAtt}_{it} +\beta_5 \text{PctMid}_{it} + \gamma_t + \varepsilon_{it}
$$

This model separates total spending from how effectively that spending translates into player performance.
