---
title: "Doms_model"
output: html_document
---

# Methods

Across all individual analyses, the goal is to evaluate how roster construction and salary allocation relate to team success. Each group member fits a distinct model, motivated by a different statistical perspective.

## Individual Analysis 1: Team-level linear regression


### Question

Does team performance increase with total payroll, and does salary concentration help or hurt after controlling for total spending?

### Model specification

```{r}
model1 <- lm(
  goals_added_dif ~ log_salary + top3_share + num_dp + num_u22 + dp_share + u22_share + strategy,
  data = team_model_data
)

broom::tidy(model1, conf.int = TRUE)
```

### Assumptions
	•	Linearity
	•	Independence across team-seasons
	•	Constant variance
	•	Limited omitted variable bias

### Why this model is appropriate

This provides a simple, interpretable baseline model linking roster spending strategy to performance.

### Model evaluation plan

	•	RMSE / MAE
	•	Residual diagnostics
	•	Compare alternative predictor sets

### Uncertainty plan

	•	95% confidence intervals for coefficients
	•	Optional bootstrap confidence intervals


## Individual Analysis 2: GAM / nonlinear team model

Initial models focused on explaining team performance (xGD) using linear regression with payroll and roster construction variables, including total salary (log_salary), salary concentration (top3_share), DP usage (dp_count or dp_salary_share), U22 usage (u22_count), and roster model classifications. Early specifications suffered from multicollinearity—particularly between top3_share, dp variables, and roster_model—and produced weak explanatory power. Re-specifying the model using xGD instead of goals added improved fit modestly, and simplifying the feature set to focus on allocation variables (especially dp_salary_share) yielded more interpretable coefficients, though statistical significance remained limited.

To test whether nonlinear effects were masking relationships, a generalized additive model (GAM) was fit with smooth terms on key variables. However, all smooth terms had effective degrees of freedom near 1, indicating no meaningful nonlinearity and confirming that a linear structure was appropriate. Given this, the final step was to introduce a single, theory-driven interaction between log_salary and dp_salary_share (Model 3). This specification tests whether the effectiveness of overall spending depends on how much of that spending is concentrated in DP-level players, aligning directly with MLS roster construction mechanisms while avoiding overfitting from excessive interactions.


## Individual Analysis 3: Player-level value model

Question

What kinds of salary allocations generate the most player value per dollar, and how does that aggregate to team-level efficiency?

Model specification

```{r, include = FALSE}
# fit_player_gam <- gam(
#   g_plus_per_90 ~ s(log(base_salary)) + position + s(log(base_salary), by = position),
#   data = player_season,
#   method = "REML"
# )
```

Aggregation step

```{r, include = FALSE}
# player_season <- player_season |>
#   mutate(predicted_value = predict(fit_player_gam, newdata = player_season))
#
# team_efficiency <- player_season |>
#   group_by(team, season) |>
#   summarize(
#     predicted_team_value = sum(predicted_value, na.rm = TRUE),
#     value_per_dollar = predicted_team_value / sum(base_salary, na.rm = TRUE),
#     .groups = "drop"
#   )
```

Why this model is appropriate

This approach directly addresses the roster construction question by first estimating player-level returns to salary, then rolling them up to the team level.

Model evaluation plan
	•	Player-level predictive performance
	•	Correlation between aggregated team value and observed team outcomes
	•	Compare linear vs nonlinear salary effects

Uncertainty plan
	•	Bootstrap the player-to-team pipeline
	•	Confidence bands by position

# Results

## Individual Analysis 1 Results

```{r, include = FALSE}
# TODO: tidy coefficient table / model summary
# broom::tidy(fit_lm, conf.int = TRUE)

# TODO: coefficient plot or fitted vs observed plot
```

We first estimated a baseline linear regression model relating team performance (goals added differential) to total payroll and salary concentration (top-three salary share).

The model explains very little variation in team performance ($R^2$ = 0.019), and neither predictor is statistically significant. While the coefficient on log payroll is positive, suggesting higher-spending teams may perform slightly better, the estimate is highly uncertain ($p$ = 0.48). Similarly, the estimated effect of salary concentration is negative but imprecise ($p$ = 0.63).

Overall, these results suggest that simple measures of payroll and salary concentration alone do not adequately explain variation in team performance in MLS.

## Individual Analysis 2 Results

```{r, include = FALSE}
# TODO: summary(fit_gam)

# TODO: plot smooth terms 
```


## Individual Analysis 3 Results

#### TODO: salary-performance curve by position

#### TODO: team value-per-dollar rankings with uncertainty bars

Key interpretation: TODO