---
title: "MLS Salary, Roster Construction, and Team Performance"
author:
  - "Dominic Hoar-Weiler (dhoarwei)"
  - "Brendan Flynn (bflynn)"
  - "Nick Kourakis (nkouraki)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: true
    toc: false
fontsize: 11pt
geometry: margin=1in
---

```{r, include = FALSE}
library(tidyverse)
library(mgcv)
library(lme4)
library(lmerTest)
library(broom)
library(broom.mixed)
library(lubridate)

# EDA objects
mls_salaries_2025 <- readRDS(
  gzcon(url("https://raw.githubusercontent.com/domihw05/FinalProject_SportsAnalyticsCourse/main/Data/mls_salaries_2025.rds"))
)

team_model_data <- readRDS(
  gzcon(url("https://raw.githubusercontent.com/domihw05/FinalProject_SportsAnalyticsCourse/main/Data/team_model_data.rds"))
)

position_salary <- readRDS(
  gzcon(url("https://raw.githubusercontent.com/domihw05/FinalProject_SportsAnalyticsCourse/main/Data/position_salary.rds"))
)

# Model objects
mls_salaries_2025 <- readRDS(
  gzcon(url("https://raw.githubusercontent.com/domihw05/FinalProject_SportsAnalyticsCourse/main/Data/mls_salaries_2025.rds"))
)

team_model_data <- readRDS(
  gzcon(url("https://raw.githubusercontent.com/domihw05/FinalProject_SportsAnalyticsCourse/main/Data/team_model_data.rds"))
)

position_salary <- readRDS(
  gzcon(url("https://raw.githubusercontent.com/domihw05/FinalProject_SportsAnalyticsCourse/main/Data/position_salary.rds"))
)

# Optional final datasets if needed for extra plots
#panel_team_data <- readRDS("Models/panel_team_data.rds")
#player_season <- readRDS("Models/player_season.rds")
#team_efficiency_model_data <- readRDS("Models/team_efficiency_model_data.rds")
#team_efficiency_model_ml_ext2 <- readRDS("Models/team_efficiency_model_ml_ext2.rds")
```

# Abstract

Major League Soccer runs under strict salary and roster rules, but teams still vary a lot in how well they perform. This project looks at how teams spend their money and whether different roster construction strategies actually lead to better results. Using publicly available MLS salary and performance data, we study how total payroll, salary distribution, and positional spending relate to outcomes like expected goal differential and points per game. We compare several statistical models that each capture a different view of roster construction. Across these approaches, we find that spending more does not guarantee success. Instead, teams that get more performance out of their salary tend to perform better. This suggests that efficiency, not just spending, is the main driver of success in MLS.

# Introduction

Major League Soccer is designed to promote parity. Teams operate under a salary cap and a set of roster rules that limit how much they can spend and how they can build their teams. Even with these constraints, there is still a large gap in performance across teams. Some teams consistently compete at the top of the table, while others struggle. This raises a simple but important question about how teams are using their limited resources. 

## Problem Statement

MLS teams all face similar financial constraints, but they make very different decisions about how to allocate salary across players and positions. There is no clear empirical answer for which of these strategies actually leads to better performance. Some teams concentrate spending on a few high-paid players, while others spread salary more evenly across the roster.

## Why This Matters

In a capped league, every dollar matters. Poor allocation decisions can limit a team’s ability to compete, while efficient spending can create an advantage even without a higher budget. From a sports analytics perspective, this is a broader question about how to allocate limited resources in a way that maximizes output. Understanding this relationship can help explain why some teams outperform others despite similar constraints.

## Research Question

How does salary allocation and roster construction relate to team performance in MLS?

More specifically, we study whether team payroll, salary concentration, and positional spending patterns are associated with outcomes such as expected goal differential, goals added, or points per game.

We use publicly available MLS salary and performance data to evaluate multiple statistical models. Each individual model addresses the same project question from a different angle, allowing us to compare which modeling framework is most informative for understanding roster construction strategy.

## Preview of Results

Our results show that higher payroll alone does not explain team success. Instead, the key factor is how efficiently teams convert salary into player performance. Teams that get more value out of their spending tend to perform better than teams that simply spend more.

# Data

We combine MLS salary information with team- and player-level performance data.

- Salary data: TODO
- Team performance data: TODO
- Player performance data: TODO
- Seasons included: TODO

## Unit of analysis

We use two related units of analysis depending on the model:

- Team-season for team-level models
- Player-season for player-level models

## Variables

### Outcome Variables

Potential response variables include:

- Team expected goal differential (xGD)
- Team goals added differential (g+ diff)
- Points per game
- Player goals added (g+)
- Minutes-adjusted player contribution

### Predictors

Potential explanatory variables include:

- Total payroll
- Log payroll
- Share of payroll spent on top 3 players
- Salary concentration index (Gini / HHI)
- Positional spending shares
- Average age / experience (if available)
- Designated player indicators (if available)

## Data Cleaning Overview

We construct a team-level dataset by aggregating player-level salary and performance data. Key steps include:

- Aggregating goals added into team-level performance
- Computing salary concentration (top 3 share)
- Creating proxies for designated players using top salary percentiles
- Constructing age-based U22 indicators
- Engineering features capturing roster balance (variance, low-salary share)

These transformations allow us to map individual salary structures to team-level outcomes.

# Exploratory Data Analysis (EDA)

## Salary distribution

```{r, echo = FALSE}
ggplot(mls_salaries_2025, aes(x = base_salary)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "MLS Salary Distribution (2025)",
       x = "Base Salary",
       y = "Count")

ggplot(mls_salaries_2025, aes(x = log(base_salary))) +
  geom_histogram(bins = 50) +
  labs(title = "Log Salary Distribution (2025)",
       x = "Log(Base Salary)",
       y = "Count")
```

## Takeaway

Salaries are highly right-skewed, meaning a small number of players earn disproportionately high salaries and suggesting a star-driven league structure.

## Team Payroll vs Team Performance

```{r, echo = FALSE, message = FALSE}
plot_data <- team_model_data

ggplot(plot_data, aes(x = log_salary, y = xgd)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Team Performance vs Payroll (2025)",
    x = "Log Total Salary",
    y = "Expected Goal Differential"
  )
```

### Takeaway

We see a weak/moderate positive relationship (likely). Payroll alone does not fully explain performance, which motivates out decision to add roster structure variables.

## Salary concentration vs team performance

```{r, echo = FALSE, message = FALSE}
concentration_plot_data <- team_model_data

ggplot(concentration_plot_data, aes(x = top3_share, y = xgd)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Salary Concentration vs Team Performance",
    x = "Top 3 Salary Share",
    y = "Expected Goals Differential"
  )
```

### Takeaway

The regression line is essentially flat. Confidence band is wide. Points are highly scattered. No visible monotonic trend.

## Positional Allocation Patterns

```{r, echo = FALSE}
ggplot(position_salary, aes(x = factor(team_id), y = pos_share, fill = position)) +
  geom_col() +
  labs(
    title = "Salary Allocation by Position (2025)",
    x = "Team",
    y = "Share of Payroll"
  ) +
  theme(axis.text.x = element_blank())
```

```{r, echo = FALSE}
ggplot(concentration_plot_data,
       aes(x = log_salary,
           y = xgd,
           size = top3_share)) +
  geom_point(alpha = 0.7) +
  labs(title = "Payroll, Concentration, and Performance",
       x = "Log Payroll",
       y = "Expected Goals Differential",
       size = "Top 3 Share")
```
\newpage

# Methods

Across all individual analyses, the goal is to evaluate how roster construction and salary allocation relate to team success. Each group member fits a distinct model, motivated by a different statistical perspective.

## Individual Analysis 1: Team-level linear regression (Dominic Hoar-Weiler)


### Question

Does team performance increase with total payroll, and does salary concentration help or hurt after controlling for total spending?

### Model specification

We model team performance using expected goal differential (xGD) at the team-season level. We begin with a simple baseline model and progressively build toward richer specifications that incorporate player-level contributions and roster construction features.

#### Baseline Model

$$
xGD_{it} = \beta_0 + \beta_1 \log(\text{Salary}_{it}) + \gamma_t + \varepsilon_{it}
$$

This model tests whether total payroll alone explains variation in team performance, controlling for season-level effects.

#### Player Level GAM

$$
GAAR_{ijt} = f(\log(\text{Salary}_{ijt})) + \delta_{\text{pos}(j)} + f_{\text{pos}(j)}(\log(\text{Salary}_{ijt})) + \varepsilon_{ijt}
$$

We estimate a nonlinear relationship between salary and player value (goals added above replacement), allowing this relationship to vary by position.

#### Team Level Construction

$$
\widehat{V}{it} = \sum_{j \in i,t} \widehat{GAAR}_{ijt} \cdot \frac{\text{Minutes}_{ijt}}{90}
$$

We aggregate predicted player contributions to construct a team-level measure of roster value.

#### Team Efficiency Model

$$
xGD_{it} = \beta_0 + \beta_1 \widehat{V}_{it} + \beta_2 \text{Salary}_{it} + \beta_3 \text{Age}_{it}
+ \beta_4 \text{PctAtt}_{it} +\beta_5 \text{PctMid}_{it} + \gamma_t + \varepsilon_{it}
$$

This model separates total spending from how effectively that spending translates into player performance.

#### Team Multilevel Model

\begin{align*}
xGD_{it} &= \beta_0 + \beta_1 \,\text{Value}_{it} + \beta_2 \,\text{Salary}_{it} + \beta_3 \,\text{Age}_{it} \\
&\quad + \beta_4 \,\text{Top3Share}_{it} + \beta_5 \,\text{U22}_{it} \\
&\quad + \beta_6 \,\text{ForwardShare}_{it} + \beta_7 \,\text{MidfieldShare}_{it} \\
&\quad + u_i + \epsilon_{it}
\end{align*}

We include team-level random effects to account for unobserved differences across teams that persist over time.

### Assumptions

Across models, we assume:

- Linearity in parameters for the regression components
- Errors satisfy $\varepsilon_{it} \sim \mathcal{N}(0, \sigma^2)$
- Independence across team-seasons (conditional on covariates and fixed/random effects)
- Constant variance of errors
- Limited omitted variable bias after including key roster and season controls

For the multilevel model, we additionally assume:

- Random effects follow $u_i \sim \mathcal{N}(0, \tau^2)$ and are independent of $\varepsilon_{it}$

### Why this model is appropriate

This modeling approach directly matches the project question by separating two effects: how much teams spend and how efficiently they convert that spending into performance. The baseline model provides a simple benchmark, while the extended models incorporate roster construction features and player-level estimates. The GAM allows for flexibility in the salary-performance relationship at the player level, which avoids imposing a strict linear assumption. The multilevel model accounts for persistent team-level differences that are not captured by observed variables.

### Model evaluation plan

We evaluate model performance using RMSE and MAE to measure predictive accuracy. We also compare models across different specifications to assess whether adding roster construction variables improves explanatory power. Residual diagnostics are used to check for violations of model assumptions, including nonlinearity and heteroskedasticity. This comparison framework allows us to determine whether more complex models provide meaningful improvements over simpler baselines.

### Uncertainty plan

We quantify uncertainty using 95% confidence intervals for all estimated coefficients. These intervals provide a measure of statistical significance and effect size. For robustness, we also consider bootstrap confidence intervals, which are less sensitive to model assumptions and provide a nonparametric check on inference. In the multilevel model, uncertainty in both fixed and random effects is accounted for through the estimated variance components.
	
### SIDE NOTES DOM

When pooling all seasons without controlling for year effects, salary variables are not statistically significant. After introducing season fixed effects, total payroll becomes significant. This suggests that league-wide changes, such as salary cap growth and evolving roster rules, confound the relationship between spending and performance if not properly controlled for.

\newpage

## Individual Analysis 2: Roster Architecture (Nick Kourakis)

### Question

Does team performance depend more on the salary disparity between highest and lowest paid starters, or on the the depth of it's TAM-eligible[^1] roster?

[^1]: TAM (Targeted Allocation Money) Players: The MLS "middle class," consisting of high-level starters earning above the Maximum Salary Budget Charge ($803,125) but below the Designated Player threshold. These players' budget hits are "bought down" using league-allocated funds. See the official documentation [here](https://www.mlssoccer.com/news/major-league-soccer-unveils-club-roster-profiles-following-2026-roster-compliance#:~:text=A%20TAM%20Player%20is%20a,are%20static%20as%20of%20Feb.).

### Model specification

#### Baseline Model

$$
xGD_{it} = \beta_0 + \beta_1 Gini_{it} + \varepsilon_{it}
$$

#### Dispersion Model

$$xGD_{it} = \beta_0 + \beta_1 \text{Gini}_{it} + \beta_2 \text{MidClassDepth}_{it} + \beta_3 \log(\text{TotalSalary}_{it}) + \varepsilon_{it}$$

Evaluates the importance of TAM players, controlling for total team spending.

#### Multi Season Roster Model

$$xGD_{it} = \beta_1 \text{Gini}_{it} + \beta_2 \text{MidClassDepth}_{it} + \alpha_i + \gamma_t + \varepsilon_{it}$$

Within a specific team over several years, does performance improve when that specific team moves toward a more balanced roster?

### Assumptions

	•	Linearity
	•	No Multicollinearity
	•	Independence


### Why this model is appropriate

Due to the MLS salary cap, every dollar spent 
on a "Star" is a dollar that could've been used somewhere else on the roster. The previous analysis 
evaluates spending efficiency, this model tests the roster structure. By using the Gini 
coefficient and TAM-depth, we can determine if success comes from a roster that is top-heavy (high inequality) or from a deep roster (high middle-class density).


### Model evaluation plan

	•	Adjusted $R^2$: To measure how much additional variance in xGD is explained 
	by roster distribution compared to a baseline model.
	•	Akaike Information Criterion (AIC): To determine if adding "Middle Class" 
	complexity improves model fit without over-fitting.
	•	Coefficient Analysis: Evaluating the significance of $\beta_1$ (Gini) to 
	identify whether inequality is correlated with performance.


### Uncertainty plan

	•	95% confidence intervals: To verify that the effects of Gini and TAM-depth are
	statistically distinguishable from zero.
	•	Clustered Standard Errors: Grouped by team_id to account for correlation in
	performance across multiple seasons.
	•	Sensitivity Analysis: Testing the model with and without extreme salary outliers. 
	
\newpage

## Individual Analysis 3 (Brendan Flynn)

### Question

### Model specification

### Assumptions

### Why this model is appropriate

### Model evaluation plan

### Uncertainty plan

\newpage

# Results

## Individual Analysis 1 Results (Dom)

Tables summarizing all model outputs are shown below. We compare models using fit ($R^2$) and predictive error, and focus on how conclusions change as we move from simple payroll measures to player-level value and multilevel structure.

```{r, eval = FALSE, echo = FALSE}
# -------------------------
# Baseline + Main LM tables
# -------------------------

baseline_tbl <- broom::tidy(panel_team_model, conf.int = TRUE)
main_tbl <- broom::tidy(team_efficiency_model, conf.int = TRUE)

knitr::kable(
  baseline_tbl,
  digits = 3,
  caption = "Baseline Team Model"
)

knitr::kable(
  main_tbl,
  digits = 3,
  caption = "Main Team Efficiency Model"
)


# -------------------------
# GAM tables
# -------------------------

gam_param_tbl <- broom::tidy(fit_player_gam, parametric = TRUE)

knitr::kable(
  gam_param_tbl,
  digits = 3,
  caption = "Player-Level GAM: Parametric Terms"
)

gam_smooth_tbl <- broom::tidy(fit_player_gam, parametric = FALSE)

knitr::kable(
  gam_smooth_tbl,
  digits = 3,
  caption = "Player-Level GAM: Smooth Terms"
)


# -------------------------
# Extended multilevel model (fixed effects)
# -------------------------

ml_fixed_tbl <- broom.mixed::tidy(
  team_efficiency_model_ml_ext2,
  effects = "fixed",
  conf.int = TRUE
) |>
  mutate(
    across(where(is.numeric), ~round(.x, 3))
  )

knitr::kable(
  ml_fixed_tbl,
  digits = 3,
  caption = "Extended Multilevel Model: Fixed Effects"
)


# -------------------------
# Extended multilevel model (random effects)
# -------------------------

ran_pars_tbl <- broom.mixed::tidy(
  team_efficiency_model_ml_ext2,
  effects = "ran_pars",
  conf.int = TRUE
) |>
  mutate(
    across(where(is.numeric), ~round(.x, 3))
  )

knitr::kable(
  ran_pars_tbl,
  digits = 3,
  caption = "Extended Multilevel Model: Random Effects"
)
```

#### Baseline Model

The baseline model explains very little variation in team performance ($R^2 \approx 0.02$), and neither total payroll nor salary concentration is statistically significant. Confidence intervals for both coefficients are wide and include zero.

This suggests that aggregate spending alone is not a meaningful predictor of team success.

#### Player Level GAM

At the player level, salary is clearly related to performance, but the relationship is nonlinear and varies by position. Higher salaries generally correspond to higher value, but with diminishing returns and different patterns across positions.

This indicates that spending does matter, but its effectiveness depends on how it is allocated across players.

#### Team Model LM

Once player contributions are aggregated to the team level, results change substantially. Predicted player value is strongly positive and statistically significant, while total salary becomes negative and significant when holding value constant.

This is the central result: teams that generate more value perform better, but conditional on value, higher spending is associated with worse outcomes. In other words, inefficient spending reduces performance.

These results are stable after adding controls such as team age and roster composition.

#### Multilevel Team Model

The multilevel model accounts for persistent differences across teams. The main findings remain unchanged: player value is strongly positive, and salary retains a negative conditional effect. Random effects indicate meaningful variation across teams, but do not alter the core relationships.

Among additional variables, age shows a negative effect, while positional spending and U22 share are not consistently significant.

Then interpretation:
	•	$\beta_1 > 0$: higher player value improves performance
	•	$\beta_2 < 0$: overpaying reduces performance
	•	$\beta_4 > 0$: top-heavy salary structures improve outcomes
	•	$\beta_5 \approx 0$: no clear U22 effect
	•	$\beta_6, \beta_7 \approx 0$: no clear positional allocation effect
	
Emphasize (these are strong)

- Value vs salary (efficiency)
- Top3 concentration effect

Neutral / exploratory

- Age

Downplay

- U22 proxy (data limitation)
- Positional shares

### Key Takeaways

- Player value is the strongest predictor of team performance
- Conditional on value, higher salary is associated with worse outcomes
- Evidence for salary concentration is positive but less consistent
- Little evidence that positional allocation or U22 spending independently drives results

### Answer to Research Question

Across all models, team performance in MLS is not driven by how much teams spend, but by how efficiently they convert salary into production.

The baseline model shows that payroll alone has little explanatory power. The player-level analysis shows that salary translates to performance, but in a nonlinear and position-dependent way. At the team level, performance is explained by total player value, while excess spending without additional value is associated with worse outcomes.

These results are consistent across model specifications and robust to team-level heterogeneity. Overall, they suggest that efficient roster construction, rather than total spending, is the key determinant of success in MLS.

## Individual Analysis 2 Results (Nick)


## Individual Analysis 3: Positional Spending (Flynn)

#### Question
  Does the allocation of salary across positions affect team performance? In particular, do teams that spend a larger share of payroll on attacking players, midfielders, defenders, or goalkeepers perform differently on average?
  
#### Model Specification
  While total payroll measures the overall financial resources a team invests into their roster, teams vary drastically in what positions they invest more heavily into. Some teams may prioritize investing heavily into forwards who are generally most responsible for increasing the expected goals of a team but also garner a higher salary. Other teams may prioritize a more balanced spending strategy, spreading spending out across positions. 
  
#### Baseline Model

$$
xGD_{it} =
\beta_0 +
\beta_1 FShare_{it} +
\beta_2 MShare_{it} +
\beta_3 DShare_{it} +
\beta_4 \log(Payroll_{it}) +
\gamma_t +
\varepsilon_{it}
$$

#### Assumptions
- Linearity
- Independence
- No Multicollinearity
- Homoskedasticity

#### Why this model is appropriate
This regression framework directly addresses the central research question by estimating the marginal effect of reallocating payroll across positions. By including total payroll in the model, we separate the overall level of spending and the distribution of spending across positions. Season fixed effects further control for league-wide changes such as rule changes, league expansion, or shifts in playing style over time. Evaluates whether teams that allocate a larger share of payroll to forwards, midfielders, or defenders perform differently on average, while controlling for season-level effects.

#### Model Evaluation Plan
Model performance will be evaluated using several metrics:

- Coefficient significance tests to determine whether positional spending shares are associated with team performance
- R-squared and adjusted R-squared to evaluate the proportion of variation in expected goal differential explained by the model
- Residual diagnostics to check for violations of model assumptions

#### Uncertainty Plan
- standard errors and confidence intervals for regression coefficients
- hypothesis testing using t-statistics and p-values
- sensitivity checks with alternative specification

### Clustering Approach

#### Model Specification
In addition to the regression model, a clustering approach is used to identify common payroll allocation strategies among MLS teams. Each team-season observation is represented by four features: Forward salary share, Midfielder salary share, Defender salary share, and Goalkeeper salary share. A k-means clustering algorithm groups teams into clusters based on these salary allocation profiles. The algorithm minimizes within-cluster variation in payroll allocation patterns.

#### Assumptions
- Teams with similar salary allocation profiles represent similar roster construction strategies.
- Euclidean distance between salary shares appropriately captures similarity between teams.
- The selected number of clusters adequately represents the underlying strategies present in the data.

#### Why this model is appropriate
While regression analysis estimates marginal effects of salary allocation, clustering identifies distinct team-building strategies present in the league. This approach allows us to examine whether teams that emphasize different positional spending strategies exhibit different performance outcomes.

#### Model Evaluation Plan
Cluster quality will be evaluated using:
- cluster sizes and stability
- visual inspection of cluster centers
- comparison of average team performance across clusters

#### Uncertainty Plan
Uncertainty in the clustering results will be addressed by:

testing alternative numbers of clusters
examining the robustness of cluster centers
comparing performance differences across clusters

```{r, include = FALSE,eval=FALSE}

# Summarize payroll by team, season, and position
pos_salary <- mls_salaries_panel %>%
  group_by(team_id, season_name, position) %>%
  summarise(
    pos_salary = sum(guaranteed_compensation, na.rm = TRUE),
    .groups = "drop"
  )

team_payroll <- mls_salaries_panel %>%
  group_by(team_id, season_name) %>%
  summarise(
    payroll = sum(guaranteed_compensation, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(season_name = as.character(season_name))

# Compute total payroll and salary shares
pos_share <- pos_salary %>%
  group_by(team_id, season_name) %>%
  mutate(
    total_salary = sum(pos_salary),
    share = pos_salary / total_salary
  ) %>%
  ungroup() %>%
  select(team_id, season_name, position, share)


# Reshape to wide format
pos_share_wide <- pos_share %>%
  pivot_wider(
    names_from = position,
    values_from = share,
    names_prefix = "",
    values_fill = 0
  ) %>%
  rename(
    FShare = F,
    MShare = M,
    DShare = D,
    GKShare = GK
  )

```

```{r, include = FALSE,eval=FALSE}
pos_share_wide2 <- pos_share_wide %>%
  mutate(season_name = as.character(season_name))

regression_data <- asa_team_xgoals_panel %>%
  select(team_id, season_name, xgoal_difference) %>%
  left_join(pos_share_wide2, by = c("team_id", "season_name"))

regression_data <- regression_data %>%
  left_join(team_payroll, by = c("team_id", "season_name"))
regression_data <- regression_data %>%
  mutate(log_payroll = log(payroll))

model2 <- lm(
  xgoal_difference ~ MShare + DShare + GKShare + log_payroll + factor(season_name),
  data = regression_data
)

summary(model2)
```

```{r, include = FALSE,eval=FALSE}
regression_data <- regression_data %>%
  mutate(
    attack_share = FShare + MShare,
    defense_share = DShare + GKShare
  )

model3 <- lm(
  xgoal_difference ~ attack_share + log_payroll + factor(season_name),
  data = regression_data
)

summary(model3)
```

```{r, include = FALSE,eval=FALSE}
#Clustered Data
cluster_data <- pos_share_wide %>%
  select(team_id, season_name, FShare, MShare, DShare, GKShare)

set.seed(2026)

clusters <- kmeans(
  cluster_data %>% select(FShare, MShare, DShare, GKShare),
  centers = 4,
  nstart = 25
)

cluster_data$cluster <- clusters$cluster

cluster_data <- cluster_data %>%
  mutate(season_name = as.character(season_name))

cluster_data <- cluster_data %>%
  left_join(
    asa_team_xgoals_panel %>%
      select(team_id, season_name, xgoal_difference),
    by = c("team_id", "season_name")
  )
```

```{r, include = FALSE, eval=FALSE}
cluster_results <- cluster_data %>%
  filter(!is.na(xgoal_difference)) %>%
  group_by(cluster) %>%
  summarise(
    avg_xGD = mean(xgoal_difference),
    sd_xGD = sd(xgoal_difference),
    teams = n()
  )

cluster_results
```

```{r, include = FALSE, eval=FALSE}
#Visualize the clusters
clusters$centers
```


```{r, include = FALSE, eval=FALSE}
library(ggplot2)

ggplot(cluster_data, aes(x = FShare, y = DShare, color = factor(cluster))) +
  geom_point(size = 2, alpha = 0.7) +
  labs(
    title = "MLS Roster Construction Clusters",
    subtitle = "Teams grouped by payroll allocation across positions",
    x = "Forward Salary Share",
    y = "Defender Salary Share",
    color = "Cluster"
  ) +
  theme_minimal()
```


```{r, include = FALSE, eval=FALSE}
cluster_centers <- as.data.frame(clusters$centers)
cluster_centers$cluster <- factor(1:nrow(cluster_centers))

cluster_centers_long <- cluster_centers %>%
  pivot_longer(-cluster, names_to = "Position", values_to = "Share")

ggplot(cluster_centers_long, aes(Position, Share, group = cluster, color = cluster)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Roster Construction Strategies in MLS",
    x = "Position",
    y = "Average Salary Share",
    color = "Cluster"
  ) +
  theme_minimal()
```


# Discussion

Our project examined whether MLS teams benefit more from spending heavily overall, concentrating spending in a few stars, or distributing spending more efficiently across the roster.

# Appendix

## Section A: Code

```{r,include = FALSE,eval=FALSE}
# Load ASA Library
library(itscalledsoccer)
library(tidyverse)
library(broom)
library(mgcv)
library(lme4)
library(tidyr)
library(lubridate)

# Instantiate a new instance of the ASA class
asa_client <- AmericanSoccerAnalysis$new()

all_games <- asa_client$games
all_teams <- asa_client$teams
all_players <- asa_client$players

asa_client$LEAGUES

# Load Roster Construction Data
mls_roster_construction <- read.csv('Data/Updated_MLS_Roster_Construction_2025.csv')
```

```{r, include = FALSE,eval=FALSE}
players <- asa_client$get_players(leagues = 'mls')
teams <- asa_client$get_teams(leagues = 'mls')

# Get team g+ information (i.e., goals added) from MLS
# tall version
asa_goals_added_tall <- asa_client$get_team_goals_added(
    leagues = "mls", season_name = 2025) |>
    tidyr::unnest(data) |>
  left_join(teams, by = c("team_id" = "team_id"))
  
asa_mls_salaries <- asa_client$get_player_salaries(leagues = 'mls') |>
  left_join(players, by = c("player_id"="player_id"))

mls_salaries_2025 <- asa_mls_salaries |>
  filter(season_name == 2025)

mls_salaries_by_team <- mls_salaries_2025 |>
  group_by(team_id) |>
  summarize(total_salary = sum(base_salary))

reference_date <- as.Date("2025-07-01")

mls_salaries_2025 <- mls_salaries_2025 |>
  mutate(
    birth_date = as.Date(birth_date),
    age = as.numeric(difftime(reference_date, birth_date, units = "days")) / 365.25
  )

```

Describe any major cleaning steps here

```{r, include = FALSE,eval=FALSE}
# -------------------------
# TEAM-LEVEL OUTCOME DATA
# -------------------------

# Team performance outcome: expected goal differential
asa_team_xgoals_tall <- asa_client$get_team_xgoals(
    leagues = "mls",
    season_name = 2025
  ) |>
  left_join(teams, by = "team_id")

team_performance <- asa_team_xgoals_tall |>
  group_by(team_id, team_name) |>
  summarize(
    xg_for = sum(xgoals_for, na.rm = TRUE),
    xg_against = sum(xgoals_against, na.rm = TRUE),
    xgd = xg_for - xg_against,
    .groups = "drop"
  )

# -------------------------
# TEAM-LEVEL SALARY FEATURES
# -------------------------

team_salary <- mls_salaries_2025 |>
  group_by(team_id) |>
  summarize(
    total_salary = sum(base_salary, na.rm = TRUE),
    roster_size = n(),
    avg_age = mean(age, na.rm = TRUE),
    .groups = "drop"
  )

salary_concentration <- mls_salaries_2025 |>
  group_by(team_id) |>
  arrange(desc(base_salary), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  summarize(
    top3_salary = sum(base_salary[rank <= 3], na.rm = TRUE),
    team_salary_tmp = sum(base_salary, na.rm = TRUE),
    top3_share = top3_salary / team_salary_tmp,
    .groups = "drop"
  ) |>
  select(team_id, top3_share)

efficiency_features <- mls_salaries_2025 |>
  group_by(team_id) |>
  summarize(
    salary_variance = var(base_salary, na.rm = TRUE),
    median_salary = median(base_salary, na.rm = TRUE),
    low_salary_share = mean(base_salary < 200000, na.rm = TRUE),
    .groups = "drop"
  )

# -------------------------
# POSITIONAL FEATURES
# -------------------------

position_salary <- mls_salaries_2025 |>
  group_by(team_id, position) |>
  summarize(
    pos_salary = sum(base_salary, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(team_id) |>
  mutate(
    team_salary = sum(pos_salary, na.rm = TRUE),
    pos_share = pos_salary / team_salary
  ) |>
  ungroup()

position_wide <- position_salary |>
  select(team_id, position, pos_share) |>
  pivot_wider(
    names_from = position,
    values_from = pos_share,
    values_fill = 0
  )
```

```{r, include = FALSE,eval=FALSE}
# -------------------------
# CLEAN ROSTER CONSTRUCTION FILE
# -------------------------

mls_roster_construction <- mls_roster_construction |>
  rename(
    team_csv = team,
    roster_model = model
  ) |>
  mutate(
    roster_model = factor(roster_model),
    strategy = case_when(
      roster_model == "3DP+3U22" ~ "DP_model",
      roster_model == "2DP+4U22" ~ "Youth_model",
      TRUE ~ "Other"
    )
  )
```

```{r, include = FALSE,eval=FALSE}
team_name_crosswalk <- tibble(
  team_csv = c(
    "Atlanta United FC",
    "Austin FC",
    "Charlotte FC",
    "Chicago Fire FC",
    "FC Cincinnati",
    "Colorado Rapids",
    "Columbus Crew",
    "FC Dallas",
    "D.C. United",
    "Houston Dynamo FC",
    "Sporting Kansas City",
    "LA Galaxy",
    "Los Angeles FC",
    "Inter Miami CF",
    "Minnesota United FC",
    "CF Montréal",
    "Nashville SC",
    "New England Revolution",
    "New York City FC",
    "New York Red Bulls",
    "Orlando City SC",
    "Philadelphia Union",
    "Portland Timbers FC",
    "Real Salt Lake",
    "San Diego FC",
    "San Jose Earthquakes",
    "Seattle Sounders FC",
    "St. Louis City SC",
    "Toronto FC",
    "Vancouver Whitecaps FC"
  ),
  team_name = c(
    "Atlanta United FC",
    "Austin FC",
    "Charlotte FC",
    "Chicago Fire FC",
    "FC Cincinnati",
    "Colorado Rapids",
    "Columbus Crew",
    "FC Dallas",
    "D.C. United",
    "Houston Dynamo FC",
    "Sporting Kansas City",
    "LA Galaxy",
    "Los Angeles FC",
    "Inter Miami CF",
    "Minnesota United FC",
    "CF Montréal",
    "Nashville SC",
    "New England Revolution",
    "New York City FC",
    "New York Red Bulls",
    "Orlando City SC",
    "Philadelphia Union",
    "Portland Timbers FC",
    "Real Salt Lake",
    "San Diego FC",
    "San Jose Earthquakes",
    "Seattle Sounders FC",
    "St. Louis City SC",
    "Toronto FC",
    "Vancouver Whitecaps FC"
  )
)
```

```{r, include = FALSE,eval=FALSE}
roster_rules_clean <- mls_roster_construction |>
  left_join(team_name_crosswalk, by = "team_csv") |>
  left_join(team_performance |> select(team_id, team_name), by = "team_name") |>
  select(team_id, roster_model, strategy, dp_count, u22_count)

team_model_data <- team_performance |>
  left_join(team_salary, by = "team_id") |>
  left_join(salary_concentration, by = "team_id") |>
  left_join(efficiency_features, by = "team_id") |>
  left_join(position_wide, by = "team_id") |>
  left_join(roster_rules_clean, by = "team_id") |>
  mutate(
    log_salary = log(total_salary),
    dp_per_roster = dp_count / roster_size,
    u22_per_roster = u22_count / roster_size
  )

# identify top dp_count players per team
dp_salary_feature <- mls_salaries_2025 |>
  group_by(team_id) |>
  arrange(desc(base_salary), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  left_join(roster_rules_clean, by = "team_id") |>
  summarize(
    dp_salary_share = sum(base_salary[rank <= dp_count], na.rm = TRUE) / sum(base_salary),
    .groups = "drop"
  )

team_model_data <- team_model_data |>
  left_join(dp_salary_feature, by = "team_id")

team_model_data |>
  select(team_id, team_name, roster_model, dp_count, u22_count) |>
  print(n = Inf)
```

```{r, include = FALSE,eval=FALSE}
# -------------------------
# MULTI-SEASON TEAM PANEL
# -------------------------

seasons <- 2013:2025

# Team xG by season
asa_team_xgoals_panel <- asa_client$get_team_xgoals(
  leagues = "mls",
  season_name = seasons,
  split_by_seasons = TRUE
) |>
  left_join(teams, by = "team_id")

team_performance_panel <- asa_team_xgoals_panel |>
  group_by(team_id, team_name, season_name) |>
  summarize(
    xg_for = sum(xgoals_for, na.rm = TRUE),
    xg_against = sum(xgoals_against, na.rm = TRUE),
    xgd = xg_for - xg_against,
    .groups = "drop"
  )

# Salary data across all seasons
mls_salaries_panel <- asa_client$get_player_salaries(
  leagues = "mls",
  season_name = seasons
) |>
  left_join(players, by = "player_id")

team_salary_panel <- mls_salaries_panel |>
  group_by(team_id, season_name) |>
  summarize(
    total_salary = sum(base_salary, na.rm = TRUE),
    roster_size = n(),
    .groups = "drop"
  )

salary_concentration_panel <- mls_salaries_panel |>
  group_by(team_id, season_name) |>
  arrange(desc(base_salary), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  summarize(
    top3_salary = sum(base_salary[rank <= 3], na.rm = TRUE),
    team_salary_tmp = sum(base_salary, na.rm = TRUE),
    top3_share = top3_salary / team_salary_tmp,
    .groups = "drop"
  ) |>
  select(team_id, season_name, top3_share)

team_performance_panel <- team_performance_panel |>
  mutate(season_name = as.integer(season_name))

team_salary_panel <- team_salary_panel |>
  mutate(season_name = as.integer(season_name))

salary_concentration_panel <- salary_concentration_panel |>
  mutate(season_name = as.integer(season_name))

panel_team_data <- team_performance_panel |>
  left_join(team_salary_panel, by = c("team_id", "season_name")) |>
  left_join(salary_concentration_panel, by = c("team_id", "season_name")) |>
  mutate(
    log_salary = log(total_salary),
    season_name = factor(season_name)
  )

panel_team_model <- lm(
  xgd ~ log_salary + top3_share + season_name,
  data = panel_team_data
)

summary(panel_team_model)
broom::tidy(panel_team_model, conf.int = TRUE)
```

```{r, include = FALSE,eval=FALSE}
# -------------------------
# PLAYER-SEASON DATA
# -------------------------

asa_player_gplus <- asa_client$get_player_goals_added(
  leagues = "mls",
  season_name = seasons,
  split_by_seasons = TRUE,
  split_by_teams = TRUE,
  above_replacement = TRUE
)

player_salaries_panel <- asa_client$get_player_salaries(
  leagues = "mls",
  season_name = seasons
) |>
  left_join(players, by = "player_id")

asa_player_gplus <- asa_player_gplus |>
  mutate(
    season_name = as.integer(season_name),
    team_id = as.character(team_id)
  )

player_salaries_panel <- player_salaries_panel |>
  mutate(
    season_name = as.integer(season_name),
    team_id = as.character(team_id)
  )

player_salaries_panel <- player_salaries_panel |>
  group_by(player_id, team_id, season_name) |>
  summarize(
    base_salary = mean(base_salary, na.rm = TRUE),
    position = first(position),
    .groups = "drop"
  )

player_season <- asa_player_gplus |>
  left_join(
    player_salaries_panel,
    by = c("player_id", "team_id", "season_name")
  ) |>
  mutate(
    log_salary = log(base_salary + 1),
    position = factor(position)
  )
```

```{r, include = FALSE,eval=FALSE}

fit_player_gam <- gam(
  goals_added_above_replacement ~ 
    s(log_salary, k = 5) +
    position +
    s(log_salary, by = position, k = 4),
  data = player_season,
  method = "REML"
)

summary(fit_player_gam)
```

```{r, include = FALSE,eval=FALSE}
player_season <- player_season |>
  mutate(
    predicted_value = predict(fit_player_gam, newdata = player_season),
    team_id = as.character(team_id),
    season_name = as.integer(season_name)
  )

player_season <- player_season |>
  left_join(players |> select(player_id, birth_date), by = "player_id") |>
  mutate(
    age = season_name - lubridate::year(birth_date)
  )

team_efficiency <- player_season |>
  group_by(team_id, season_name) |>
  summarize(
    predicted_team_value = sum(predicted_value * minutes_played / 90, na.rm = TRUE),
    player_total_salary = sum(base_salary, na.rm = TRUE),
    roster_sz = n_distinct(player_id),
    pct_att = mean(position == "F"),
    pct_midfielders = mean(position == "M"),
    avg_age = mean(age, na.rm = TRUE),
    .groups = "drop"
  )

panel_team_data <- panel_team_data |>
  mutate(
    team_id = as.character(team_id),
    season_name = as.integer(as.character(season_name))
  )

team_efficiency_model_data <- panel_team_data |>
  left_join(team_efficiency, by = c("team_id", "season_name")) |>
  filter(!is.na(xgd), !is.na(predicted_team_value), !is.na(player_total_salary))
```

```{r, include = FALSE,eval=FALSE}
team_efficiency_model <- lm(
  xgd ~ predicted_team_value + player_total_salary +  factor(season_name) +
    avg_age + pct_att + pct_midfielders,
  data = team_efficiency_model_data
)


summary(team_efficiency_model)
broom::tidy(team_efficiency_model, conf.int = TRUE)
```

```{r, include = FALSE,eval=FALSE}
team_efficiency_model_data <- team_efficiency_model_data |>
  mutate(
    pred_val_s = scale(predicted_team_value),
    salary_s = scale(player_total_salary),
    age_s = scale(avg_age),
    pct_att_s = scale(pct_att),
    pct_mid_s = scale(pct_midfielders)
  )

team_efficiency_model_ml <- lmer(
  xgd ~ pred_val_s + salary_s + age_s + pct_att_s + pct_mid_s +
    (1 | team_id) + (1 | season_name),
  data = team_efficiency_model_data
)

summary(team_efficiency_model_ml)

library(lmerTest)

model_no_team <- lmer(
  xgd ~ pred_val_s + salary_s + age_s + pct_att_s + pct_mid_s +
    (1 | season_name),
  data = team_efficiency_model_data
)

anova(model_no_team, team_efficiency_model_ml)
```
