---
title: "MLS Salary, Roster Construction, and Team Performance"
author:
  - "Dominic Hoar-Weiler (dhoarwei)"
  - "Brendan Flynn (bflynn)"
  - "Nick Kourakis (nkouraki)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: true
    toc: false
fontsize: 11pt
geometry: margin=1in
---

# Introduction

Major League Soccer operates under a complex roster-building system, yet teams vary widely in on-field performance. 

## Problem Statement

MLS teams operate under strict roster and salary constraints, yet there is no clear empirical consensus on which roster construction strategies lead to success.

## Why This Matters

Understanding how teams should allocate limited salary budget is critical for competitive strategy in a capped league like MLS.

## Research Question

How does salary allocation and roster construction relate to team performance in MLS?

More specifically, we study whether team payroll, salary concentration, and positional spending patterns are associated with outcomes such as expected goal differential, goals added, or points per game.

We use publicly available MLS salary and performance data to evaluate multiple statistical models. Each individual model addresses the same project question from a different angle, allowing us to compare which modeling framework is most informative for understanding roster construction strategy.

## Preview of Results

NEED TO ADD

# Data

We combine MLS salary information with team- and player-level performance data.
	•	Salary data: TODO
	•	Team performance data: TODO
	•	Player performance data: TODO
	•	Seasons included: TODO

## Unit of analysis

We use two related units of analysis depending on the model:
	•	Team-season for team-level models
	•	Player-season for player-level models

## Variables

#### Outcome Variables

Potential response variables include:
	•	Team expected goal differential (xGD)
	•	Team goals added differential (g+ diff)
	•	Points per game
	•	Player goals added (g+)
	•	Minutes-adjusted player contribution

#### Predictors

Potential explanatory variables include:
	•	Total payroll
	•	Log payroll
	•	Share of payroll spent on top 3 players
	•	Salary concentration index (Gini / HHI)
	•	Positional spending shares
	•	Average age / experience (if available)
	•	Designated player indicators (if available)

## Data Cleaning Overview

We construct a team-level dataset by aggregating player-level salary and performance data. Key steps include:

- Aggregating goals added into team-level performance
- Computing salary concentration (top 3 share)
- Creating proxies for designated players using top salary percentiles
- Constructing age-based U22 indicators
- Engineering features capturing roster balance (variance, low-salary share)

These transformations allow us to map individual salary structures to team-level outcomes.


# Exploratory Data Analysis (EDA)

### Salary distribution

```{r, echo = FALSE}
ggplot(mls_salaries_2025, aes(x = base_salary)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "MLS Salary Distribution (2025)",
       x = "Base Salary",
       y = "Count")

ggplot(mls_salaries_2025, aes(x = log(base_salary))) +
  geom_histogram(bins = 50) +
  labs(title = "Log Salary Distribution (2025)",
       x = "Log(Base Salary)",
       y = "Count")
```

#### Takeaway

Salaries are highly right-skewed, meaning a small number of players earn disproportionately high salaries and suggesting a star-driven league structure.

### Team Payroll vs Team Performance

```{r, echo = FALSE}
plot_data <- team_model_data

ggplot(plot_data, aes(x = log_salary, y = goals_added_dif)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Team Performance vs Payroll (2025)",
    x = "Log Total Salary",
    y = "Goals Added Differential"
  )
```

#### Takeaway

We see a weak/moderate positive relationship (likely). Payroll alone does not fully explain performance, which motivates out decision to add roster structure variables.

### Salary concentration vs team performance

```{r, echo = FALSE}
concentration_plot_data <- team_model_data

ggplot(concentration_plot_data, aes(x = top3_share, y = goals_added_dif)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Salary Concentration vs Team Performance",
    x = "Top 3 Salary Share",
    y = "Goals Added Differential"
  )
```

#### Takeaway

The regression line is essentially flat. Confidence band is wide. Points are highly scattered. No visible monotonic trend.

### Positional Allocation Patterns

```{r, echo = FALSE}
ggplot(position_salary, aes(x = factor(team_id), y = pos_share, fill = position)) +
  geom_col() +
  labs(
    title = "Salary Allocation by Position (2025)",
    x = "Team",
    y = "Share of Payroll"
  ) +
  theme(axis.text.x = element_blank())
```

EDA takeaway: TODO

```{r, echo = FALSE}
ggplot(concentration_plot_data,
       aes(x = log_salary,
           y = goals_added_dif,
           size = top3_share)) +
  geom_point(alpha = 0.7) +
  labs(title = "Payroll, Concentration, and Performance",
       x = "Log Payroll",
       y = "Goals Added Differential",
       size = "Top 3 Share")
```
\newpage

# Methods

Across all individual analyses, the goal is to evaluate how roster construction and salary allocation relate to team success. Each group member fits a distinct model, motivated by a different statistical perspective.

## Individual Analysis 1: Team-level linear regression (Dominic Hoar-Weiler)


### Question

Does team performance increase with total payroll, and does salary concentration help or hurt after controlling for total spending?

### Model specification

```{r}
model1 <- lm(
  xgd ~ log_salary + dp_salary_share + u22_count +
        log_salary:dp_salary_share,
  data = team_model_data
)

summary(model1)

broom::tidy(model1, conf.int = TRUE)
```

```{r}
gam1 <- gam(
  xgd ~ s(log_salary, k = 4) + s(dp_salary_share, k = 4) + s(top3_share,k=4) + u22_count ,
  data = team_model_data,
  method = "REML"
)

summary(gam1)

plot(gam1,pages = 1, residuals = TRUE)
gam.check(gam1)
```

```{r}
player_season <- player_season |>
  mutate(predicted_value = predict(fit_player_gam, newdata = player_season))

team_efficiency <- player_season |>
  group_by(team_id) |>
  summarize(
    predicted_team_value = sum(predicted_value, na.rm = TRUE),
    total_salary = sum(base_salary, na.rm = TRUE),
    value_per_dollar = predicted_team_value / total_salary,
    .groups = "drop"
  )
```

### Assumptions
	•	Linearity
	•	Independence across team-seasons
	•	Constant variance
	•	Limited omitted variable bias

### Why this model is appropriate

This provides a simple, interpretable baseline model linking roster spending strategy to performance.

### Model evaluation plan

	•	RMSE / MAE
	•	Residual diagnostics
	•	Compare alternative predictor sets

### Uncertainty plan

	•	95% confidence intervals for coefficients
	•	Optional bootstrap confidence intervals
	
### SIDE NOTES DOM

When pooling all seasons without controlling for year effects, salary variables are not statistically significant. However, once season fixed effects are introduced, total payroll becomes a significant predictor of team performance. This suggests that cross-season structural changes in MLS—such as salary cap growth and evolving roster rules—obscure the relationship between spending and performance when not properly controlled for.

\newpage

## Individual Analysis 2 (Nick Kourakis)

### Question

### Model specification

### Assumptions

### Why this model is appropriate

### Model evaluation plan

### Uncertainty plan

\newpage

## Individual Analysis 3 (Brendan Flynn)

### Question

### Model specification

### Assumptions

### Why this model is appropriate

### Model evaluation plan

### Uncertainty plan

\newpage

# Results

## Individual Analysis 1 Results (Dom)

```{r, include = FALSE}
# TODO: tidy coefficient table / model summary
# broom::tidy(fit_lm, conf.int = TRUE)

# TODO: coefficient plot or fitted vs observed plot
```

We first estimated a baseline linear regression model relating team performance (goals added differential) to total payroll and salary concentration (top-three salary share).

The model explains very little variation in team performance ($R^2$ = 0.019), and neither predictor is statistically significant. While the coefficient on log payroll is positive, suggesting higher-spending teams may perform slightly better, the estimate is highly uncertain ($p$ = 0.48). Similarly, the estimated effect of salary concentration is negative but imprecise ($p$ = 0.63).

Overall, these results suggest that simple measures of payroll and salary concentration alone do not adequately explain variation in team performance in MLS.

------

At the player level, the results show a clear but nuanced relationship between salary and performance. Higher salaries are associated with greater goals added above replacement, but the effect is distinctly nonlinear and varies by position. Attackers and midfielders, in particular, generate more value on average, and the marginal returns to salary differ across roles. This suggests that while spending more does generally translate into better individual performance, the returns are not uniform and begin to diminish, especially depending on how that money is allocated across positions.

At the team level, once we properly aggregate player contributions using minutes-weighted predicted value, a much clearer structure emerges. Teams with higher total player value perform significantly better, as expected. However, controlling for this value, total salary has a negative and significant effect on performance. In other words, spending more does not directly lead to better results—what matters is how efficiently that spending is converted into on-field production. Together, these results point to a consistent takeaway: salary matters because it helps acquire talent, but team success ultimately depends on how effectively that talent translates into actual contributions during play.

At the player level, salary clearly translates into performance, but not in a simple linear way. The GAM shows strong nonlinear effects and meaningful differences across positions, with attackers and midfielders generating more value and different salary-to-performance curves depending on role. So spending does matter, but how that money is allocated across players and positions is just as important as how much is spent.

At the team level, once player contributions are properly aggregated using minutes-weighted predicted value, performance becomes much more predictable. Teams with higher total player value perform significantly better, while total salary—holding value constant—has a negative effect. This indicates that inefficiency in spending, rather than spending itself, is what hurts performance. Adding team-level controls like age and roster composition does not change this core result. Predicted player value remains strongly positive and significant, and salary remains negatively associated with performance, reinforcing the idea that converting payroll into actual on-field production is what drives success.

One additional insight is that team age matters: older teams tend to perform worse, even after accounting for value and spending. In contrast, roster size and positional composition do not have strong independent effects once player value is accounted for. Overall, the results consistently point to the same conclusion: success is not about how much teams spend, but how efficiently they turn that spending into real contributions on the field.

## Individual Analysis 2 Results (Nick)


## Individual Analysis 3 Results (Flynn)



# Discussion

Our project examined whether MLS teams benefit more from spending heavily overall, concentrating spending in a few stars, or distributing spending more efficiently across the roster.

# Appendix

## Section A: Code

```{r, include = FALSE}
# Load ASA Library
library(itscalledsoccer)
library(tidyverse)
library(broom)
library(mgcv)
library(lme4)
library(tidyr)
library(lubridate)

# Instantiate a new instance of the ASA class
asa_client <- AmericanSoccerAnalysis$new()

all_games <- asa_client$games
all_teams <- asa_client$teams
all_players <- asa_client$players

asa_client$LEAGUES

# Load Roster Construction Data
mls_roster_construction <- read.csv('Updated_MLS_Roster_Construction_2025.csv')
```

```{r}
players <- asa_client$get_players(leagues = 'mls')
teams <- asa_client$get_teams(leagues = 'mls')

# Get team g+ information (i.e., goals added) from MLS
# tall version
asa_goals_added_tall <- asa_client$get_team_goals_added(
    leagues = "mls", season_name = 2025) |>
    tidyr::unnest(data) |>
  left_join(teams, by = c("team_id" = "team_id"))
  
asa_mls_salaries <- asa_client$get_player_salaries(leagues = 'mls') |>
  left_join(players, by = c("player_id"="player_id"))

mls_salaries_2025 <- asa_mls_salaries |>
  filter(season_name == 2025)

mls_salaries_by_team <- mls_salaries_2025 |>
  group_by(team_id) |>
  summarize(total_salary = sum(base_salary))

reference_date <- as.Date("2025-07-01")

mls_salaries_2025 <- mls_salaries_2025 |>
  mutate(
    birth_date = as.Date(birth_date),
    age = as.numeric(difftime(reference_date, birth_date, units = "days")) / 365.25
  )

```

Describe any major cleaning steps here

```{r}
# -------------------------
# TEAM-LEVEL OUTCOME DATA
# -------------------------

# Team performance outcome: expected goal differential
asa_team_xgoals_tall <- asa_client$get_team_xgoals(
    leagues = "mls",
    season_name = 2025
  ) |>
  left_join(teams, by = "team_id")

team_performance <- asa_team_xgoals_tall |>
  group_by(team_id, team_name) |>
  summarize(
    xg_for = sum(xgoals_for, na.rm = TRUE),
    xg_against = sum(xgoals_against, na.rm = TRUE),
    xgd = xg_for - xg_against,
    .groups = "drop"
  )

# -------------------------
# TEAM-LEVEL SALARY FEATURES
# -------------------------

team_salary <- mls_salaries_2025 |>
  group_by(team_id) |>
  summarize(
    total_salary = sum(base_salary, na.rm = TRUE),
    roster_size = n(),
    avg_age = mean(age, na.rm = TRUE),
    .groups = "drop"
  )

salary_concentration <- mls_salaries_2025 |>
  group_by(team_id) |>
  arrange(desc(base_salary), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  summarize(
    top3_salary = sum(base_salary[rank <= 3], na.rm = TRUE),
    team_salary_tmp = sum(base_salary, na.rm = TRUE),
    top3_share = top3_salary / team_salary_tmp,
    .groups = "drop"
  ) |>
  select(team_id, top3_share)

efficiency_features <- mls_salaries_2025 |>
  group_by(team_id) |>
  summarize(
    salary_variance = var(base_salary, na.rm = TRUE),
    median_salary = median(base_salary, na.rm = TRUE),
    low_salary_share = mean(base_salary < 200000, na.rm = TRUE),
    .groups = "drop"
  )

# identify top dp_count players per team
dp_salary_feature <- mls_salaries_2025 |>
  group_by(team_id) |>
  arrange(desc(base_salary), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  left_join(roster_rules_clean, by = "team_id") |>
  summarize(
    dp_salary_share = sum(base_salary[rank <= dp_count], na.rm = TRUE) / sum(base_salary),
    .groups = "drop"
  )

# -------------------------
# POSITIONAL FEATURES
# -------------------------

position_salary <- mls_salaries_2025 |>
  group_by(team_id, position) |>
  summarize(
    pos_salary = sum(base_salary, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(team_id) |>
  mutate(
    team_salary = sum(pos_salary, na.rm = TRUE),
    pos_share = pos_salary / team_salary
  ) |>
  ungroup()

position_wide <- position_salary |>
  select(team_id, position, pos_share) |>
  pivot_wider(
    names_from = position,
    values_from = pos_share,
    values_fill = 0
  )
```

```{r}
# -------------------------
# CLEAN ROSTER CONSTRUCTION FILE
# -------------------------

mls_roster_construction <- mls_roster_construction |>
  rename(
    team_csv = team,
    roster_model = model
  ) |>
  mutate(
    roster_model = factor(roster_model),
    strategy = case_when(
      roster_model == "3DP+3U22" ~ "DP_model",
      roster_model == "2DP+4U22" ~ "Youth_model",
      TRUE ~ "Other"
    )
  )
```

```{r}
team_name_crosswalk <- tibble(
  team_csv = c(
    "Atlanta United FC",
    "Austin FC",
    "Charlotte FC",
    "Chicago Fire FC",
    "FC Cincinnati",
    "Colorado Rapids",
    "Columbus Crew",
    "FC Dallas",
    "D.C. United",
    "Houston Dynamo FC",
    "Sporting Kansas City",
    "LA Galaxy",
    "Los Angeles FC",
    "Inter Miami CF",
    "Minnesota United FC",
    "CF Montréal",
    "Nashville SC",
    "New England Revolution",
    "New York City FC",
    "New York Red Bulls",
    "Orlando City SC",
    "Philadelphia Union",
    "Portland Timbers FC",
    "Real Salt Lake",
    "San Diego FC",
    "San Jose Earthquakes",
    "Seattle Sounders FC",
    "St. Louis City SC",
    "Toronto FC",
    "Vancouver Whitecaps FC"
  ),
  team_name = c(
    "Atlanta United FC",
    "Austin FC",
    "Charlotte FC",
    "Chicago Fire FC",
    "FC Cincinnati",
    "Colorado Rapids",
    "Columbus Crew",
    "FC Dallas",
    "D.C. United",
    "Houston Dynamo FC",
    "Sporting Kansas City",
    "LA Galaxy",
    "Los Angeles FC",
    "Inter Miami CF",
    "Minnesota United FC",
    "CF Montréal",
    "Nashville SC",
    "New England Revolution",
    "New York City FC",
    "New York Red Bulls",
    "Orlando City SC",
    "Philadelphia Union",
    "Portland Timbers FC",
    "Real Salt Lake",
    "San Diego FC",
    "San Jose Earthquakes",
    "Seattle Sounders FC",
    "St. Louis City SC",
    "Toronto FC",
    "Vancouver Whitecaps FC"
  )
)
```

```{r}
roster_rules_clean <- mls_roster_construction |>
  left_join(team_name_crosswalk, by = "team_csv") |>
  left_join(team_performance |> select(team_id, team_name), by = "team_name") |>
  select(team_id, roster_model, strategy, dp_count, u22_count)

team_model_data <- team_performance |>
  left_join(team_salary, by = "team_id") |>
  left_join(salary_concentration, by = "team_id") |>
  left_join(efficiency_features, by = "team_id") |>
  left_join(position_wide, by = "team_id") |>
  left_join(roster_rules_clean, by = "team_id") |>
  mutate(
    log_salary = log(total_salary),
    dp_per_roster = dp_count / roster_size,
    u22_per_roster = u22_count / roster_size
  )

team_model_data <- team_model_data |>
  left_join(dp_salary_feature, by = "team_id")

team_model_data |>
  select(team_id, team_name, roster_model, dp_count, u22_count) |>
  print(n = Inf)
```

```{r}
# -------------------------
# MULTI-SEASON TEAM PANEL
# -------------------------

seasons <- 2015:2025

# Team xG by season
asa_team_xgoals_panel <- asa_client$get_team_xgoals(
  leagues = "mls",
  season_name = seasons,
  split_by_seasons = TRUE
) |>
  left_join(teams, by = "team_id")

team_performance_panel <- asa_team_xgoals_panel |>
  group_by(team_id, team_name, season_name) |>
  summarize(
    xg_for = sum(xgoals_for, na.rm = TRUE),
    xg_against = sum(xgoals_against, na.rm = TRUE),
    xgd = xg_for - xg_against,
    .groups = "drop"
  )

# Salary data across all seasons
mls_salaries_panel <- asa_client$get_player_salaries(
  leagues = "mls",
  season_name = seasons
) |>
  left_join(players, by = "player_id")

team_salary_panel <- mls_salaries_panel |>
  group_by(team_id, season_name) |>
  summarize(
    total_salary = sum(base_salary, na.rm = TRUE),
    roster_size = n(),
    .groups = "drop"
  )

salary_concentration_panel <- mls_salaries_panel |>
  group_by(team_id, season_name) |>
  arrange(desc(base_salary), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  summarize(
    top3_salary = sum(base_salary[rank <= 3], na.rm = TRUE),
    team_salary_tmp = sum(base_salary, na.rm = TRUE),
    top3_share = top3_salary / team_salary_tmp,
    .groups = "drop"
  ) |>
  select(team_id, season_name, top3_share)

team_performance_panel <- team_performance_panel |>
  mutate(season_name = as.integer(season_name))

team_salary_panel <- team_salary_panel |>
  mutate(season_name = as.integer(season_name))

salary_concentration_panel <- salary_concentration_panel |>
  mutate(season_name = as.integer(season_name))

panel_team_data <- team_performance_panel |>
  left_join(team_salary_panel, by = c("team_id", "season_name")) |>
  left_join(salary_concentration_panel, by = c("team_id", "season_name")) |>
  mutate(
    log_salary = log(total_salary),
    season_name = factor(season_name)
  )

panel_team_model <- lm(
  xgd ~ log_salary + top3_share + season_name,
  data = panel_team_data
)

summary(panel_team_model)
broom::tidy(panel_team_model, conf.int = TRUE)
```

```{r}
# -------------------------
# PLAYER-SEASON DATA
# -------------------------

asa_player_gplus <- asa_client$get_player_goals_added(
  leagues = "mls",
  season_name = seasons,
  split_by_seasons = TRUE,
  split_by_teams = TRUE,
  above_replacement = TRUE
)

player_salaries_panel <- asa_client$get_player_salaries(
  leagues = "mls",
  season_name = seasons
) |>
  left_join(players, by = "player_id")

asa_player_gplus <- asa_player_gplus |>
  mutate(
    season_name = as.integer(season_name),
    team_id = as.character(team_id)
  )

player_salaries_panel <- player_salaries_panel |>
  mutate(
    season_name = as.integer(season_name),
    team_id = as.character(team_id)
  )

player_salaries_panel <- player_salaries_panel |>
  group_by(player_id, team_id, season_name) |>
  summarize(
    base_salary = mean(base_salary, na.rm = TRUE),
    position = first(position),
    .groups = "drop"
  )

player_season <- asa_player_gplus |>
  left_join(
    player_salaries_panel,
    by = c("player_id", "team_id", "season_name")
  ) |>
  mutate(
    log_salary = log(base_salary + 1),
    position = factor(position)
  )
```

```{r}

fit_player_gam <- gam(
  goals_added_above_replacement ~ 
    s(log_salary, k = 5) +
    position +
    s(log_salary, by = position, k = 4),
  data = player_season,
  method = "REML"
)

summary(fit_player_gam)
```

```{r}
player_season <- player_season |>
  mutate(
    predicted_value = predict(fit_player_gam, newdata = player_season),
    team_id = as.character(team_id),
    season_name = as.integer(season_name)
  )

player_season <- player_season |>
  left_join(players |> select(player_id, birth_date), by = "player_id") |>
  mutate(
    age = season_name - lubridate::year(birth_date)
  )

team_efficiency <- player_season |>
  group_by(team_id, season_name) |>
  summarize(
    predicted_team_value = sum(predicted_value * minutes_played / 90, na.rm = TRUE),
    player_total_salary = sum(base_salary, na.rm = TRUE),
    roster_sz = n_distinct(player_id),
    pct_att = mean(position == "F"),
    pct_midfielders = mean(position == "M"),
    avg_age = mean(age, na.rm = TRUE),
    .groups = "drop"
  )

panel_team_data <- panel_team_data |>
  mutate(
    team_id = as.character(team_id),
    season_name = as.integer(as.character(season_name))
  )

team_efficiency_model_data <- panel_team_data |>
  left_join(team_efficiency, by = c("team_id", "season_name")) |>
  filter(!is.na(xgd), !is.na(predicted_team_value), !is.na(player_total_salary))
```

```{r}
team_efficiency_model <- lm(
  xgd ~ predicted_team_value + player_total_salary +  factor(season_name) +
    avg_age + pct_att + pct_midfielders,
  data = team_efficiency_model_data
)

summary(team_efficiency_model)
broom::tidy(team_efficiency_model, conf.int = TRUE)
```
